---
title: "OHIO T1DM 2020 Pre-Processing"
author: ""
date: "2025-01-02"
output: pdf_document
---

```{r config, include=FALSE}
# Source configuration
config_locations <- c(
  file.path(getwd(), "cma_cluster", "config.R"),
  file.path(getwd(), "..", "cma_cluster", "config.R"),
  file.path(getwd(), "..", "..", "cma_cluster", "config.R")
)
for (cfg_path in config_locations) {
  if (file.exists(cfg_path)) {
    source(cfg_path)
    break
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required libraries
library(dplyr)  # For bind_rows in 5-min processing
```

# OhioT1DM 2020 Dataset Preprocessing

This script processes the 2020 OhioT1DM dataset, which contains 6 additional subjects:
- Subject IDs: 540, 544, 552, 567, 584, 596

The processing pipeline mirrors the 2018 preprocessing but includes additional
robustness for potential XML structure differences between cohorts.

```{r}
#-----------------------------------------------
# Data preprocessing
#-----------------------------------------------
source("realdata-preprocess-functions.R")

# Helper function to inspect XML structure
inspect_xml_structure <- function(xml_path) {
  data <- xmlParse(xml_path)
  rootnode <- xmlRoot(data)
  variables <- xmlChildren(rootnode)
  varnames <- sapply(variables, xmlName)
  cat("Variables found:", paste(varnames, collapse = ", "), "\n")
  return(varnames)
}
```

```{r}
#---------------------------------
# 1. Extract data from xml - Training data

data_path <- CONFIG$OHIO_2020_TRAIN_DIR

# 2020 cohort subject IDs
file_names_train <- c("540-ws-training.xml",
                      "544-ws-training.xml",
                      "552-ws-training.xml",
                      "567-ws-training.xml",
                      "584-ws-training.xml",
                      "596-ws-training.xml")

nperson <- length(file_names_train)

# Inspect XML structure from first file to understand 2020 format
cat("\n=== Inspecting 2020 XML Structure ===\n")
varnames_2020 <- inspect_xml_structure(file.path(data_path, file_names_train[1]))

raw_train <- list()
varnames_list <- list()  # Store varnames for each subject

for (n in 1:nperson) {
  data <- xmlParse(paste(data_path, file_names_train[n], sep = "/"))
  rootnode <- xmlRoot(data)
  variables <- xmlChildren(rootnode)
  nvar <- xmlSize(rootnode)
  raw_train[[n]] <- list()
  varnames <- rep("", nvar)
  for (i in 1:nvar) {
    vari <- variables[[i]]
    varnames[i] <- xmlName(variables[[i]])
    data_var <- t(as.matrix(xmlSApply(vari, xmlAttrs)))
    raw_train[[n]][[i]] <- data_var
  }
  varnames_list[[n]] <- varnames  # Store for later use
  cat(sprintf("Loaded training data for subject %d: %s (%d variables)\n",
              n, file_names_train[n], nvar))
}

# Store the first subject's varnames as the reference
varnames_2020 <- varnames_list[[1]]
cat("\n2020 Variable names:", paste(varnames_2020, collapse = ", "), "\n")
```

```{r}
#---------------------------------
# 2. Extract data from xml - Testing data

data_path <- CONFIG$OHIO_2020_TEST_DIR

file_names_test <- c("540-ws-testing.xml",
                     "544-ws-testing.xml",
                     "552-ws-testing.xml",
                     "567-ws-testing.xml",
                     "584-ws-testing.xml",
                     "596-ws-testing.xml")

raw_test <- list()
varnames_list_test <- list()  # Store varnames for each subject

for (n in 1:nperson) {
  data <- xmlParse(paste(data_path, file_names_test[n], sep = "/"))
  rootnode <- xmlRoot(data)
  variables <- xmlChildren(rootnode)
  nvar <- xmlSize(rootnode)
  raw_test[[n]] <- list()
  varnames <- rep("", nvar)
  for (i in 1:nvar) {
    vari <- variables[[i]]
    varnames[i] <- xmlName(variables[[i]])
    data_var <- t(as.matrix(xmlSApply(vari, xmlAttrs)))
    raw_test[[n]][[i]] <- data_var
  }
  varnames_list_test[[n]] <- varnames
  cat(sprintf("Loaded testing data for subject %d: %s (%d variables)\n",
              n, file_names_test[n], nvar))
}
```

#-----------------------------------------------
# 3. Preprocess data into 30-minute intervals

```{r}
interval <- 30  # separate by half an hour
data_train <- list()
data_test  <- list()
data_full  <- list()

for (n in 1:nperson) {
  cat(sprintf("\n=== Processing subject %d ===\n", n))

  # training data - use preprocess0_2020 for 2020 data format
  train0 <- preprocess0_2020(raw_train[[n]], varnames = varnames_list[[n]])
  train1 <- preprocess1(train0, interval_min = interval)
  data_train[[n]] <- preprocess2(train1, interval_min = interval)

  # testing data
  startdate <- date(dmy_hms(train0$glucose[1, 1]))
  startbasal <- data_train[[n]]$basal[length(data_train[[n]]$basal)]
  test0 <- preprocess0_2020(raw_test[[n]], varnames = varnames_list_test[[n]])
  test1 <- preprocess1(test0, interval_min = interval, startdate = startdate)
  data_test[[n]] <- preprocess2(test1, interval_min = interval, startbasal = startbasal)

  # combine training and testing (use bind_rows to handle column mismatches)
  data_full[[n]] <- bind_rows(data_train[[n]], data_test[[n]])

  # generate extra variables
  # total amount of carb input in the next half an hour
  data_full[[n]] <- preprocess3_future(data_full[[n]],
                                       variable = "meal",
                                       interval_min = interval,
                                       lag_time_range = c(0, 30),
                                       new_var_name = "meal_future_halfhour",
                                       FUN = sum)

  # total amount of bolus from last time interval
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "bolus",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "bolus_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "steps",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "steps_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "heart",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "avg_heart_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "gsr",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "avg_gsr_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "skintemp",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "avg_skintemp_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "airtemp",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "avg_airtemp_past_30mins",
                                     FUN = sum)

  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "meal",
                                     interval_min = interval,
                                     lag_time_range = c(30, 60),
                                     new_var_name = "meal_past_30mins",
                                     FUN = sum)

  # average basal rate of 4-8 hours before the current time point
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "basal",
                                     interval_min = interval,
                                     lag_time_range = c(240, 480),
                                     new_var_name = "basal_4_8")

  # total amount of carb input in the previous 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "meal",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "meal_past_2hrs",
                                     FUN = sum)

  # total amount of steps from past 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "steps",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "steps_past_2hrs",
                                     FUN = sum)

  # total bolus taken at the meal plus the next 4 intervals
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "bolus",
                                     interval_min = interval,
                                     lag_time_range = c(30, 180),
                                     new_var_name = "bolus_past_2hrs",
                                     FUN = sum)

  # average heart rate in past 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "heart",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "avg_heart_past_2hrs")

  # average gsr in past 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "gsr",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "avg_gsr_past_2hrs")

  # average skintemp in past 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "skintemp",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "avg_skintemp_past_2hrs")

  # average airtemp in past 2 hours
  data_full[[n]] <- preprocess3_past(data_full[[n]],
                                     variable = "airtemp",
                                     interval_min = interval,
                                     lag_time_range = c(30, 150),
                                     new_var_name = "avg_airtemp_past_2hrs")
}

save.image("data_preprocessed_2020-halfhour.RData")
```

#--------
# 4. Preprocess into 5-minute intervals
#--------

```{r}
interval <- 5   # now 5-minute bins
data_train <- vector("list", nperson)
data_test  <- vector("list", nperson)
data_full  <- vector("list", nperson)

for (n in seq_len(nperson)) {
  cat(sprintf("\n=== Processing subject %d (5-min intervals) ===\n", n))

  # training data - use preprocess0_2020 for 2020 data format
  train0 <- preprocess0_2020(raw_train[[n]], varnames = varnames_list[[n]])
  train1 <- preprocess1(train0, interval_min = interval)
  data_train[[n]] <- preprocess2(train1, interval_min = interval)

  # testing data
  startdate <- date(dmy_hms(train0$glucose[1, 1]))
  startbasal <- data_train[[n]]$basal[length(data_train[[n]]$basal)]
  test0 <- preprocess0_2020(raw_test[[n]], varnames = varnames_list_test[[n]])
  test1 <- preprocess1(test0, interval_min = interval, startdate = startdate)
  data_test[[n]] <- preprocess2(test1, interval_min = interval, startbasal = startbasal)

  # combine
  data_full[[n]] <- bind_rows(data_train[[n]], data_test[[n]])

  # future carb in next 5-min window
  data_full[[n]] <- preprocess3_future(
    data_full[[n]],
    variable       = "meal",
    interval_min   = interval,
    lag_time_range = c(0, interval),
    new_var_name   = "meal_future_5min",
    FUN            = sum
  )

  # past-window summaries (all lag ranges unchanged, but interval_min=5)
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "bolus", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "bolus_past_5_10min", FUN = sum
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "steps", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "steps_past_5_10min", FUN = sum
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "heart", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "avg_heart_past_5_10min"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "gsr", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "avg_gsr_past_5_10min"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "skintemp", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "avg_skintemp_past_5_10min"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "airtemp", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "avg_airtemp_past_5_10min"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "meal", interval_min = interval,
    lag_time_range = c(5, 10),   new_var_name = "meal_past_5_10min", FUN = sum
  )

  # basal averaged 4-8 hrs before (4h=240min, 8h=480min)
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "basal", interval_min = interval,
    lag_time_range = c(240, 480), new_var_name = "basal_4_8"
  )

  # carb past 2 hrs (30-150 min)
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "meal", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "meal_past_2hrs", FUN = sum
  )
  # steps past 2 hrs
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "steps", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "steps_past_2hrs", FUN = sum
  )
  # bolus past 2 hrs (30-180 min)
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "bolus", interval_min = interval,
    lag_time_range = c(30, 180),  new_var_name = "bolus_past_2hrs", FUN = sum
  )
  # average heart, gsr, skintemp, airtemp past 2 hrs
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "heart", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "avg_heart_past_2hrs"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "gsr", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "avg_gsr_past_2hrs"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "skintemp", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "avg_skintemp_past_2hrs"
  )
  data_full[[n]] <- preprocess3_past(
    data_full[[n]], "airtemp", interval_min = interval,
    lag_time_range = c(30, 150),  new_var_name = "avg_airtemp_past_2hrs"
  )
}

# Save the 5-min processed 2020 data
save.image("data_preprocessed_2020-5min.RData")

cat("\n=== 2020 Data Preprocessing Complete ===\n")
cat(sprintf("Processed %d subjects\n", nperson))
cat(sprintf("Interval: %d minutes\n", interval))
```

#--------------------------------------------
# 5. Plot part of the 5min dataset for QC
#--------------------------------------------

```{r}
load("data_preprocessed_2020-5min.RData")

i <- 1  # First subject
dataset <- data_full[[i]]
startdate <- 7
enddate <- 13
nday <- enddate - startdate + 1
ntt <- 60 * 24 / 5
temp <- subset(dataset, (day >= startdate) & (day <= enddate))
temp$timestamp <- (temp$day - startdate) * ntt + temp$tt

# plot the glucose
par(mar = c(5, 4, 4, 6) + 0.1)
plot(temp$timestamp, temp$glucose,
     ylim = c(0, 400),
     xlim = c(0, ntt * nday), type = "l", axes = FALSE,
     xlab = "", ylab = "",
     main = paste("2020 Cohort - Mobile Health Data for Subject", i, "in", nday, "days"))
labels <- rep("", 2 * nday + 1)
labels[(1:nday) * 2] <- paste("Day ", c(1 : nday))
axis(1, at = (ntt / 2 * c(0:(nday * 2))), labels = labels, col = "black")
abline(h = 80, col = "green")
abline(h = 140, col = "green")
mtext("Timestamp with 5 Min Intervals", side = 1, line = 2.5)
box()
par(new = TRUE)

# plot the meals
temp2 <- subset(temp, meal > 0)
plot(temp2$timestamp, temp2$meal,
     xlim = c(0, ntt * nday), pch = 20,
     xlab = "", ylab = "", ylim = c(0, 200),
     axes = FALSE, col = "red", col.axis = "red")
arrows(temp2$timestamp, 0, temp2$timestamp, temp2$meal, length = 0, col = "red")
mtext("Carb Intake (g)", side = 4, col = "red", line = 2.5)
axis(4, ylim = c(0, 200), col = "red", col.ticks = "red")

# plot insulin dosage
par(new = TRUE)
temp3 <- subset(temp, bolus > 0)
plot(temp3$timestamp, temp3$bolus,
     xlim = c(0, ntt * nday), pch = 20,
     xlab = "", ylab = "", col = "blue", ylim = c(0, 20), axes = FALSE)
arrows(temp3$timestamp, 0, temp3$timestamp, temp3$bolus, length = 0, col = "blue")
axis(2, ylim = c(0, 20), col = "blue", col.ticks = "blue")
mtext("Insulin Dosage (Unit)", side = 2, col = "blue", line = 2.5)

legend("topleft", col = c("black", "green"),
       lty = c(1, 1), legend = c("Blood Glucose (mg/dL)", "Safe Blood Glucose Range"))
```
